Simulation Study of Angular Resolution of a New Electromagnetic Sampling Calorimeter

### 2022-02-23 paper.tex
updates

\begin{abstract}
We report simulation results on the angular resolution of electromagnetic (EM) sampling calorimeters with photons being in the range of 100~MeV to 2~GeV. A simulation model of the EM calorimeter consists of alternating layers of a 1-mm-thick Pb plate and a 5-mm-thick plastic scintillator plate. The scintillator plates are alternatively segmented in horizontal and vertical strips. In this study, we obtained energy deposit in individual strips using Geant4 simulations and reconstructed incident photon angles using a XGBoost model with boosted decision trees. Angular resolutions are studied in terms of the strip width. The 15-mm-wide strips provide the best angular resolution of 1.23~$\pm$~0.01$^{\circ}$. Incident photon directions can be well reconstructed in the front part (5$X_{0}$) of the EM calorimeter still possessing the acceptable angular resolution of 1.33~$\pm$~0.01$^{\circ}$. The angular resolution is almost constant for small angles less than 30$^{\circ}$, and varies significantly with the incident photon energy. The
energy dependence can be expressed as $\sigma_{\theta}=$~1.21~$\oplus$~0.45~$/\sqrt{E_{\gamma}}$.

@@

\section{Motivation}
\label{sec:mot}
The electromagnetic (EM) calorimeter plays an important role in experimental studies for the nuclear and particle physics. Various materials have been developed for better energy and timing resolutions so far~\cite{Calorimeter}. EM calorimeters can be divided into two groups: sampling calorimeters and homogeneous calorimeter. Sampling calorimeters consist of alternating layers of an absorber generating EM showers and an active medium providing signals. On the other hand, homogeneous calorimeters are built of only a single type of material that facilitates both EM showers and signal generations. Today, the sampling calorimeter becomes very popular in large-scale high energy experiments, owing to its cost-effectiveness. In sampling calorimeters the energy deposited in the active medium fluctuates because the active layers are interleaved with absorber layers. This so-called sampling fluctuation dominates the energy resolution.

The sampling fluctuation can be improved by optimizing the thickness ratio of the absorber to the active medium. A 3-m-long cylindrical calorimeter made of alternating layers of Pb and plastic scintillating plates is an example of the sampling calorimeter~\cite{Murayama:2020mcp}.

@@

This paper describes a machine learning approach based on XGBoost (XGB) model~\cite{xgboost:2016} with boosted decision trees in order to reconstruct the directions of incident photons entering a sampling calorimeter with alternating layers of a Pb plate and plastic scintillator strips in horizontal and vertical directions. We generated EM showers in a sampling calorimeter using Geant4 simulations. We report on the detector configuration providing the acceptable angular resolution, and report on the angular resolution of the EM sampling calorimeter in terms of the incident angle and energy.

@@

Simulation models of a sampling calorimeter were constructed as a block consisting of alternating layers of a 1-mm-thick Pb absorber and a 5-mm-thick polyvinyltoluene-based plastic scintillator. The plastic scintillator is segmented in 15-mm-wide strips, which is alternatively oriented in vertical and horizontal directions as shown in Fig.~\ref{fig:det_conf}. The sampling calorimeter model has a cross-section of 525~$\times$~525~mm$^{2}$ and accommodates 105 alternating layers of 630~mm length (20$X_{0}$), which is sufficiently long to absorb full photon energy in the range of 100~MeV to 2~GeV.

@@

Incidence angles of photons are reconstructed using the XGB model that is a scalable machine learning system for tree boosting~\cite{xgboost:2016}. A machine learning model maps a set of data inputs, known as features, to target variables. In this study, the XGB model maps a dataset of energy deposits in each scintillator strip to an incidence angle of a photon. Training data were carefully prepared using Geant4 simulation such that the input datasets are representative of a detector response of real sampling calorimeters. To minimize 
data bias, the incidence angles were uniformly generated at the detector surface in the angular range of 0~$<\theta<$~50$^{\circ}$ and 0~$<\varphi<$~360$^{\circ}$, where $\varphi$ denotes azimuthal angle. Such a wide angular coverage provides good training datasets for the XGB. The number of training samples is $10^{5}$ considering limited computing resources. To test the reconstruction of incidence angles, we generated photons at a fixed incidence angle $\theta$. Note that the incident energy is known.

@@

We studied a correlation between byperparameters of the XGB model and the angular resolution. The test scans the evaluated angular resolution with different hyperparameters. Figure~\ref{fig:par_scan} shows the test results with N\_estimators and Max. depth. N\_estimators defines the maximum allowed number of decision trees to be developed, and the Max. depth defines the complexity of the structure of decision trees. The best hyperparameter combination was searched for in terms of the angular resolution. As a results, N\_estimators and Max. depth are set to 300 and 100, respectively. Similar tests were also performed for different hyperparameters. The best values for other hyperparameters are displayed in Tab.~\ref{tab:XgbPar}. Other hyperparameters tuned in this study are Subsample, Learning rate, and Gamma. Subsample controls the fraction of total event samples for each boosting procedure, and Learning rate weights a decision tree to be added current model. Lastly, Gamma regulates the evaluation of each decision tree.

\begin{table}[hbt!]{\small
\centering
\caption{Hyperparameters of the $\XGB$}
\begin{tabular}{cccc}
\hline 
Parameter & Function & Default value & Used value \\ \hline 
N\_estimators & The number of decision trees & N.A. & 300 \\  
Max. depth & Possible maximum depth of tree structure & 6 & 100 \\ 
Subsample & Fraction of total data used for a single decision & 1 & 0.8 \\ 
Learning rate & Step length for calculation & 0.3 & 0.02 \\ 
Gamma & Requirement on minimum loss function & 0 & 0 \\ 
\hline
\end{tabular}
\label{tab:XgbPar}
}\end{table}

@@

Furthermore, incidence angles are reconstructed with the front layers of $X_{0}$ in which 99\% of incident photons generate EM showers. Figure~\ref{fig:angle_reco_layer} (a) shows reconstructed angles using front 24 layers of the detector, which corresponds to 4.6$X_{0}$ for 1-GeV photons. A fraction of photons failed to be reconstructed due to the lack of active layers. The failed events are represented as a delta function near 0, and such events are removed by requiring the total energy deposit to be larger than 10~MeV. The angular resolution with the front layer is estimated to be 1.33~$\pm$~0.01$^{\circ}$ while possessing inefficiency estimated to be 5.8\%.

@@

Figure~\ref{fig:angle_reco_dep_gr} (a) shows the angular resolution as a function of the incident photon energy ($E_{\gamma}=$) for $\theta=$~10$^{\circ}$. Angular resolutions are fitted with a function of $p_{0} \oplus p_{1}/\sqrt{E_{\gamma}(GeV)}$, where $p_{0}$ represents an energy-independent contribution and is estimated to be 0.45$^{\circ}$, and $p_{1}$ denotes the energy-dependent parameter, mainly related with the development of the EM shower. $p_{1}$ is estimated to be xxx$^{\circ}$ for $\theta>$~yyy$^{\circ}$ and xxx$^{\circ}$ for $\theta=$~yyy$^{\circ}$ in Fig.~\ref{fig:angle_reco_dep_gr} (b). The discrepancy between them is attributed to the followings: polar angle to be always positive and poor angular resolutions for low $E_{\gamma}$ photons. When $\mu$ and $\alpha$ get comparable each other in Eq~\ref{eqn:gg}, the left tail of the distribution would extend to negative area. This largely affects the case $\mu=$~10$^{\circ}$ and not the case $\mu>$~15$^{\circ}$ considering that the angular resolutions for $E_{\gamma}=$~0.1~GeV is estimated to be 4$^{\circ}$.

@@

Figure~\ref{fig:multi-parameter} shows the angular resolution with different numbers of training samples for different detector configurations. All configurations show a decrease in the angular resolution with increasing training samples, which means that enhanced statistics for the training results in better reconstruction. The angular resolution with 5-mm-wide scintillator strips decreases more rapidly than others. The number of features for the 5-mm-wide strip is much larger than others, and the number of training samples is much required to correlate larger features. Considering limited computing resources, the setup with $10^{5}$ events, 24 layers, and the 15-mm-wide strip was chosen for training the $\XGB$ model. The training result differs by only 0.1$^{\circ}$ from the best setup will full layers.

@@

We report simulation results on the incidece angle reconstruction of EM sampling calorimeters with photons being in the range of 100~MeV to 2~GeV. EM shwers are simulated in the new samplinng calorimeter.using Geant4. We utilize the $\XGB$ model to reconstruct the photon incident angle by correlating all energy deposits in each strip with the incidence angle all events together.

We tune the hyperparameter of the $\XGB$ to reconstruct the incidence angle with the best angular resolution, and study the angular resolution in terms of the detector configuration. We find that 15-mm-wide strips provide the best angular resolution, which can be expressed as 0.45~$\oplus~$1.21$/\sqrt{E_\gamma}$ for different incident angles. We also conclude that using 24 front layers gives the angular resolution which is comparable with the angular resolution we get from the full layer.

We study that the angular resolution is changing with the variation of the training of the $\XGB$ and the detector configuration. As the strip width decreases, more channels of scintillator strips are required, and the training of the $\XGB$ gets affected due to the increased feature size. This is quenched with the larger number of training samples, which qualitatively improves the training of the $\XGB$. On the other hand, the longer strip width provides the worse angular resolution as the position resolution of the EM shower gets worse. We conclude that 15-mm-wide strips provide the favorable angular resolution with 24 front layers and $10^{5}$ training samples.

>>>>

\begin{abstract}
We report simulation results on the angular resolution of electromagnetic (EM) sampling calorimeters with photons being in the range of 100~MeV to 2~GeV. A simulation model of the EM calorimeter consists of alternating layers of a 1-mm-thick Pb plate and a 5-mm-thick plastic scintillator plate. The scintillator plates are alternatively segmented in horizontal and vertical strips. In this study, we obtained energy deposits in individual strips using Geant4 simulations and reconstructed incident photon angles using a XGBoost model with boosted decision trees. Angular resolutions are studied in terms of the strip width. The 15-mm-wide strips provide the best angular resolution of 1.23~$\pm$~0.01$^{\circ}$. Incident photon directions can be well reconstructed in the front part (5$X_{0}$) of the EM calorimeter with the acceptable angular resolution of 1.33~$\pm$~0.01$^{\circ}$. The angular resolution is mildly dependent on the incidence angle, and varies significantly with the incident photon energy. The energy dependence can be expressed as $\sigma_{\theta}=$~0.43~$\oplus p1 /\sqrt{E_{\gamma}}$, where $p1$ gradually decreases with the incidence angle.

@@

\section{Motivation}
\label{sec:mot}
Electromagnetic (EM) calorimeter plays an important role in experimental studies for the nuclear and particle physics. Various materials have been developed for better energy and timing resolutions so far~\cite{Calorimeter}. EM calorimeters can be divided into two groups: sampling calorimeters and homogeneous calorimeters. Sampling calorimeters consist of alternating layers of an absorber generating EM showers and an active medium providing signals. On the other hand, homogeneous calorimeters are built of only a single type of material that facilitates both EM showers and signal generations. Today, the sampling calorimeter becomes very popular in large-scale high energy experiments, owing to its cost-effectiveness. In sampling calorimeters the energy deposited in the active medium fluctuates because the active layers are interleaved with absorber layers. This so-called sampling fluctuation dominates the energy resolution.

The sampling fluctuation can be improved by optimizing the thickness ratio of the absorber to the active medium. A 3-m-long cylindrical calorimeter made of alternating layers of Pb and plastic scintillator plates is an example of the sampling calorimeter~\cite{Murayama:2020mcp}.

@@

This paper describes a machine learning approach based on XGBoost (XGB) model~\cite{xgboost:2016} with boosted decision trees in order to reconstruct the direction of incident photons entering a sampling calorimeter with alternating layers of a Pb plate and plastic scintillator strips in horizontal and vertical directions. We generated EM showers in a sampling calorimeter using Geant4 simulations. We report on the detector configuration providing the acceptable angular resolution, and report on the angular resolution of the EM sampling calorimeter in terms of the incident angle and energy.

@@

Simulation models of a sampling calorimeter were constructed as a block consisting of alternating layers of a 1-mm-thick Pb absorber and a 5-mm-thick polyvinyltoluene-based plastic scintillator. The plastic scintillator is segmented in 15-mm-wide strips, which are alternatively oriented in vertical and horizontal directions as shown in Fig.~\ref{fig:det_conf}. The sampling calorimeter model has a cross-section of 525~$\times$~525~mm$^{2}$ and accommodates 105 alternating layers of 630~mm length (20$X_{0}$), which is sufficiently long to absorb full photon energy in the range of 100~MeV to 2~GeV.

@@

Incidence angles of photons are reconstructed using the XGB model that is a scalable machine learning system for the gradient boosting tree~\cite{xgboost:2016}. A machine learning model maps a set of data inputs, known as features, to target variables. In this study, the XGB model maps a dataset of energy deposits in each scintillator strip to an incidence angle of a photon. Training data were carefully prepared using Geant4 simulation such that the input datasets are representative of a detector response of real sampling calorimeters. To minimize 
data bias, the incidence angles were uniformly generated at the detector surface in the angular range of 0~$<\theta<$~50$^{\circ}$ and 0~$<\varphi<$~360$^{\circ}$, where $\varphi$ denotes azimuthal angle. Such a wide angular coverage provides good training datasets for the XGB. The number of training samples is $10^{5}$ considering limited computing resources. To test the reconstruction of incidence angles, we generated photons at a fixed incidence angle $\theta$. Note that the incidence energy is known.

@@

We studied a correlation between byperparameters of the XGB model and the angular resolution. The test scans the evaluated angular resolution with different hyperparameters. Figure~\ref{fig:par_scan} shows the test results with N\_estimators and Max. depth. N\_estimators defines the maximum allowed number of decision trees to be developed, and the Max. depth defines the complexity of the structure of decision trees. The best hyperparameter combination was searched for in terms of the angular resolution. As a results, N\_estimators and Max. depth are set to 300 and 100, respectively. Similar tests were also performed for different hyperparameters. The best values for other hyperparameters are displayed in Tab.~\ref{tab:XgbPar}. Other hyperparameters tuned in this study are Subsample, Learning rate, and Gamma. Subsample controls the fraction of total event samples for each boosting procedure, and Learning rate weights a decision tree to be added onto the current model. Lastly, Gamma regulates the evaluation of each decision tree.

\begin{table}[hbt!]{\small
\centering
\caption{Hyperparameters of the XGB model}
\begin{tabular}{cccc}
\hline 
Parameter & Function & Default value & Used value \\ \hline 
N\_estimators & The number of decision trees & N.A. & 300 \\  
Max. depth & Possible maximum depth of tree structure & 6 & 100 \\ 
Subsample & Fraction of total data used for a single decision & 1 & 0.8 \\ 
Learning rate & Step length for calculation & 0.3 & 0.02 \\ 
Gamma & Requirement on minimum loss function & 0 & 0 \\ 
\hline
\end{tabular}
\label{tab:XgbPar}
}\end{table}

@@

Furthermore, incidence angles are reconstructed with the front layers of $X_{0}$ in which 99\% of incident photons generate EM showers. Figure~\ref{fig:angle_reco_layer} (a) shows reconstructed angles using front 24 layers of the detector, which correspond to 4.6$X_{0}$ for 1-GeV photons. A fraction of photons failed to be reconstructed due to the lack of active layers. The failed events are represented as a delta function near 0, and such events are removed by requiring the total energy deposit to be larger than 10~MeV, 1\% of incidence energy. The angular resolution with the front layer is estimated to be 1.33~$\pm$~0.01$^{\circ}$ while possessing inefficiency estimated to be 5.8\%.

@@

Figure~\ref{fig:angle_reco_dep_gr} (a) shows the angular resolution as a function of the incident photon energy ($E_{\gamma}$) at $\theta=$~10$^{\circ}$. Angular resolutions are fitted with a function of $p_{0} \oplus p_{1}/\sqrt{E_{\gamma}(GeV)}$, where $p_{0}$ represents an energy-independent contribution and is estimated to be 0.45$^{\circ}$, and $p_{1}$ denotes the energy-dependent parameter, mainly related with the development of the EM shower. $p_{1}$ is estimated to be xxx$^{\circ}$ for $\theta>$~yyy$^{\circ}$ and xxx$^{\circ}$ for $\theta=$~yyy$^{\circ}$ in Fig.~\ref{fig:angle_reco_dep_gr} (b). The discrepancy between them is attributed to the followings: polar angle to be always positive and poor angular resolutions for low $E_{\gamma}$ photons. When $\mu$ and $\alpha$ get comparable each other in Eq.~\ref{eqn:gg}, the left tail of the distribution would extend to negative area. This largely affects the case $\mu=$~10$^{\circ}$ and not the case $\mu>$~15$^{\circ}$ considering that the angular resolutions for $E_{\gamma}=$~0.1~GeV is estimated to be 4$^{\circ}$.

@@

Figure~\ref{fig:multi-parameter} shows the angular resolution with different numbers of training samples for different detector configurations. All configurations show a decrease in the angular resolution with increasing training samples, which means that enhanced statistics for the training results in better reconstruction. The angular resolution with 5-mm-wide scintillator strips decreases more rapidly than others. The number of features for the 5-mm-wide strip is much larger than others, and the number of training samples is much required to correlate larger features. Considering limited computing resources, the setup with $10^{5}$ events, 24 layers, and the 15-mm-wide strip was chosen for training the XGB model. The training result differs by only 0.2$^{\circ}$ from the best setup will full layers.

@@

We report simulation results on the incidece angle reconstruction of EM sampling calorimeters with photons being in the range of 100~MeV to 2~GeV. EM shwers are simulated in the new samplinng calorimeter.using Geant4. We utilize the XGB model to reconstruct the photon incident angle by correlating all energy deposits in each strip with the incidence angle all events together.

We tune the hyperparameter of the XGB to reconstruct the incidence angle with the best angular resolution, and study the angular resolution in terms of the detector configuration. We find that 15-mm-wide strips provide the best angular resolution, which can be expressed as 0.45~$\oplus~$xxx$/\sqrt{E_\gamma}$ for different incidence angles. We also conclude that using 24 front layers gives the angular resolution which is comparable with the angular resolution we get from the full layer.

Furthermore, we study that the angular resolution is changing with the variation of the training of the XGB and the detector configuration. As the strip width decreases, more channels of scintillator strips are required, and the training of the XGB gets affected due to the increased feature size. This is quenched with the larger number of training samples, which qualitatively improves the training of the XGB. On the other hand, the longer strip width provides the worse angular resolution as the position resolution of the EM shower gets worse. We conclude that 15-mm-wide strips provide the favorable angular resolution with 24 front layers and $10^{5}$ training samples.
